Visual narratives of the Covid-19 pandemic

### 2022-01-22 JDSSV_manuscript_knitr.Rnw
Additional information on comparisons and misinformation as well as some tweaks to the english in the first couple of sections

At the same time we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.

@@

Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge.

While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. Thus, in studying COVID-19 and predicting its spread, it is not only important to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results. Associations observed in the data can often be caused by third party variables (confounders). In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.

\section[]{The global perspective}

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global perspective of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example Figure~\ref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. For the pandemic perspective a central element of the narrative is the ubiquity of the disease and the accompanying global impact. Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.  
\end{enumerate}

@@

Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage.

Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region.

@@

And clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often exercies a western-centric viewpoint.

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe. None of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. The use of maps in this context more often focusses on the comparative aspects: how are we doing as opposed to our neighbors, which startegy to fight Covid-19 is better? Topics that we will look closer into in Section~\ref{sec:rankings}

@@

While traditionally, time-series information has been limited to charts with time on the x-axis, there have been many attempts to make use of the capabilities of web graphics and animation in order to convey temporal information in less traditional time-series forms. One controversial chart appeared in the New York Times; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}. The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}. The re-envisioned version, which aligns counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons. In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly. Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period.

@@

% bar-chart race https://www.youtube.com/watch?v=8WH-bTH6GqI

@@

\section[]{Comparisons and rankings}
\label{sec:rankings}

The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}. Sociologists have coined the term "othering" for such attributions of the others \citep{mountz2009}. This refers to the observation that above all new, unknown threats are projected onto "strangers" and "the others". Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease. Unsurprisingly, the prevailing visual narrative focused on comparisons often fuelled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data at all types of adminsitratory levels, fostered ongoing competition and the derivation of leaderboards.

@@

The bar charts used for this purpose (see top image in Fig.~\ref{fig:ft-death-ranking}) are straigthforward and provide typically a good way to show the "leading" countries or regions. Emphasis on specific countries is either preconfigured by the creator of the chart or can be modified interactively in the online version by check boxes and drop-down menus. These bar charts also allow easy adjustment of the count data to population size or other meaningful standard. Streamgraphs (see lower image in Fig.~\ref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008, being used by the New York Times to visualise box office results of movies \citep{streamgraph}. Their usefulness depends heavily on the clarity of the pattern. Various design choices such as the ordering of the different groups have a strong impact on the visual impression and message conveyed by the graph. Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time.

% https://twitter.com/AndToddsaid/status/1314231055096971265 is a horrible set of graphs and a great example of the snark associated with covid management policy and geographic comparisons...

>>>>

At the same time, we still see a lot of defective graphics disseminated and shared: some that violate fundamental visualisation and statistical reporting principles such as accuracy, relevance, timeliness, clarity, coherence, and reproducibility. These principles have been laid out in numerous standards for statistical reporting in the application areas, such as the ESS standard for quality reports \citep{ess2009}, the CONSORT, PRISMA, CHEERs guidelines, and others (see https://equator-network.org). Numerous publications, initiatives, and ideas to improve the communication of quantitative and statistical information have been prepared, see for example \cite{Hoffrage2261,Tufte2001,Rosling2011,otavamylona2020}.

@@

Nevertheless, both the complexity of phenomena and the "bipolarity" of statistical thinking remain a challenge. While human thinking tends to simplify patterns and political communication also prefers a simple cause-effect relationship, real phenomena are often multivariate. Thus, in studying COVID-19 and predicting its spread, it is not only important to consider symptomatology, disease incidence and geographic distribution, population behaviour patterns, government policies, and impacts on the economy, schools, people in nursing homes, and society at large, but also to incorporate these into data analyses and communication of results. Associations observed in the data can often be caused by third party variables (confounders). In addition, much of the data comes from observational studies, which usually makes robust causal attribution problematic. However, statisticians who point out these limitations are at risk of having their statements pulled out one-sidedly in a polarised debate \citep{McConway2021}.

\section{The global perspective}

On 11 March 2020, WHO declared the outbreak of the novel coronavirus disease (COVID-19) a pandemic, and since that date at the latest, the global distribution of the disease has been in the public eye. The spatial spread of the virus and the resulting cases and deaths are commonly visualized by choropleth maps, see for example \Cref{fig:choro1} showing the total number of infections reported in each country as of January 14, 2022. For the pandemic perspective, a central element of the narrative is the ubiquity of the disease and the accompanying global impact. Choropleth maps based on raw numbers of cases might look convincing and fit to the purpose, but neglect a number of well-know caveats for statistical reporting and visualisation:
\begin{enumerate}
\item Absolute value unsuitability: As explained in \citep{monmonier2005, slocum2008, speckmann2010} among others, choropleth maps are fundamentally unsuitable for the representation of absolute numbers. Especially in the case of similarly coloured areas of the regions, viewers tend to integrate them unconsciously and perceive choropleths as representations of density. They also do not help to convey the desired message as the absolute numbers of Covid-19 cases are strongly influenced by the population size of the country, but also by the number of tests performed and the accuracy of the recording and reporting system.
\item The area-bias: The visual impression is determined more by the colour and the geographical area of the individual countries than by the number of Covid cases. Since the countries of the world differ extremely in area, the visual assessment is distorted, especially in the case of neighbouring countries with similar numbers but different areas.
\item Color-scheme obstructions: Much research in visualisation is concerned with the appropriate choice of colour schemes, see \citep{brewer1997}. The choice of a continuous scale or a categorical scale, the choice of a scale that promotes the recognition of patterns or a scale that supports the filtering out of specific map details, influences the quality of choropleth maps.  
\end{enumerate}

@@

Choropleth maps are quite commonly used by media companies and governmental organisations and it is easy to find good and bad examples of their usage. Proportional symbol maps, see Fig.~\ref{fig:propsymb} or graduated symbol maps place scaled symbols or diagrams directly on the input map, often on the centroid of the regions. The symbol, most commonly a disk or a square, is scaled such that its area corresponds to the data value of the region.

@@

And clearly, the visual impression of any map based illustration depends on the projection chosen for the underlying map as the examples quite often demonstrate a western-centric viewpoint.

The narrative of the global perspective seems to be highly limited to the aspect of a pandemic affecting the entire globe. None of the above viusalisations intends to provide a deeper insight into the spatial distribution of the phenomenon, neither within the administratively motivated spatial borders nor across them. The use of maps in this context more often focuses on the comparative aspects: how are we doing as opposed to our neighbors, which strategy to fight Covid-19 is better? We will examine these topics more closely in Section~\ref{sec:rankings}.

@@

While traditionally, time-series information has been limited to charts with time on the x-axis, there have been many attempts to make use of the capabilities of web graphics and animation in order to convey temporal information in less traditional time-series forms. One controversial chart appeared in the New York Times; the x-axis was wrapped around an Archimedean spiral to provide a sense of periodicity in the year-by-year evolution of case counts, as shown in \Cref{fig:nyt-spiral}. The original form of this chart is prone to the line-width illusion \citep{vanderplasSignsSineIllusion2015}, in addition to the well-known problems with polar charts \citep{hofmannGraphicalTestsPower2012,waldnerComparisonRadialLinear2020}. The re-envisioned version, which aligns counts on the outside of the spiral, and displays them as radial lines, mitigates some of the issues with the line-width illusion by directly showing each line as its own entity, facilitating direct comparisons. In addition, the re-envisioned chart includes a reference line at 100K cases/day, which allows the reader to compare the severity of different waves directly. Still, it is more work to interpret and compare peaks on this chart than on a similar linear time series chart with the same data, as the reader must assess line width and then do a mental rotation and shift operation in order to compare with any other time period. In addition, this type of chart devotes less area to previous years of the pandemic, which may decrease the amount of focus given to that data; whether this is an advantage or disadvantage depends on your perspective and goals when using the chart.

@@

Another attempt to creatively show time-series data that has been extremely common online throughout the pandemic is the bar chart race: an animated series of bar charts over time that show how case counts in each country are increasing. A typical example can be found on YouTube at \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}; a screenshot is shown in \Cref{fig:barchart-race}. This type of chart is effective in showing instantaneous changes in counts over days for countries with an extremely high number of cases, but may make it difficult for the viewer to process how cases are changing in countries that do not dominate case counts. By focusing viewer attention on changes in relative totals of case counts, the graphic tends to hide steady, proportional increases in cases across the globe, making it somewhat difficult to perceive large global trends relative to small changes in comparative case counts.

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures_Web/Bar-chart-race-screenshot}
\caption{A screenshot from Stats on Clock's bar chart race showing counts of covid cases over time throughout the pandemic. Source: \url{https://www.youtube.com/watch?v=8WH-bTH6GqI}}\label{fig:barchart-race}
\end{figure}

@@

\section{Comparisons and rankings}
\label{sec:rankings}

The plague is always the others: This is the short formula for dealing with infectious diseases from a historical perspective \citep{thiessen2021}. Sociologists have coined the term ``othering'' for such attributions of the others \citep{mountz2009}. This refers to the observation that above all new, unknown threats are projected onto ``strangers'' and ``the others''. Closing the borders and restricting access to the country became a popular means of controlling the spread of the disease at several different points during the pandemic. Unsurprisingly, the prevailing visual narrative focused on comparisons often fueled by political rivalry, historical dependencies, recent withdrawal from supranational institutions or regional competitions. An ongoing debate about the true extent of the dangers of covid-19 and how best to combat it, combined with daily availability and public access to data across administrative levels, fostered ongoing competition and the use of leaderboards to show how bad case counts were somewhere else.

@@

The bar charts used for this purpose (see top image in Fig.~\ref{fig:ft-death-ranking}) are straightforward and provide a good way to show the ``leading'' countries or regions. Emphasis on specific countries is either preconfigured by the creator of the chart or can be modified interactively in the online version by check boxes and drop-down menus. These bar charts also allow easy adjustment of the count data to population size or other meaningful standard. Streamgraphs (see lower image in Fig.~\ref{fig:ft-death-ranking}) are a special type of stacked area graphs and became quite popular around 2008 when they were used by the New York Times to visualise box office results of movies \citep{streamgraph}. Their usefulness depends heavily on the clarity of the pattern. Various design choices such as the ordering of the different groups have a strong impact on the visual impression and message conveyed by the graph. Small variations in the proportions of the groups are almost impossible to detect, but streamgraphs give a good indication of which group is predominant at any given time, even though they are subject to the line-width or sine illusion \citep{vanderplasSignsSineIllusion2015}.

Often, these comparative charts were combined with policy discussions, with individuals challenged to spot certain policy interventions on the case-count graphs of different localities with different policies\citep{weissThese12Graphs2020}. \Cref{fig:policy-comparison} shows a chart from an article which appeared in The Federalist, a conservative outlet in the United States, suggesting that mask mandates are ineffective because it is difficult to spot the impact of the mask mandate on the overall trend of cases. Of course, mask mandates are but one component of a much broader pandemic management strategy, but the goal of the chart is clear: the reader is supposed to conclude that cases and masks are not associated. These charts were so common on social media that media outlets ran stories identifying them as misinformation \citep{reutersstaffFactCheckMask2020}; clearly, comparative charts were effectively deployed to misinform as well as to inform.

\begin{figure}
\centering
\includegraphics[width=.8\linewidth]{Figures_Web/Deaths-NY-TX-GA-Sweden-Federalist.jpeg}
\caption{Deaths per million people in New York, Sweden, Texas, and Georgia, as shown in the Federalist \citep{weissThese12Graphs2020}. The goal of this comparison chart is to lead the reader to conclude that masks are ineffective, as there is not a clear difference between Texas and Georgia. New York is also shown (though not labeled as requiring or not requiring masks), and Sweden is shown as well, despite not being a US state at all, leading the critical reader to conclude that the time series shown in this chart have been cherry-picked to make a rhetorical point.}\label{fig:policy-comparison}
\end{figure}
